Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-01-25 NovelLBM.tex
Redo Nozzle

\title{Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids}
\author{Richard Kellnberger$^1$}
\date{October, 2023}
\begin{document}
	\maketitle

@@

	We need this solution to compare to literature. However, this solution does not include $\eta_\text{s}$, which is required in general. With the inclusion of the solvent viscosity an analytical solution can no longer be found and the use of a semi-analytic solution is required (for details see appendix \ref{app:PTT}).
	In the following we compare results for channel and pipe flow to these equations for realistic parameters. Afterward we compare the accuracy of our simulation algorithm against literature.
	%We use richard/vtk 4dbc4d379842c538f084e68a5dcfe2555a674495 (2024-01-09)
	\subsection{Channel flow vs theory}
	Channel flow (also known as 2D Poiseuille) is one of the simplest flows and a very typical example. We pick a channel with a radius of \SI{10}{\micro\meter}, meaning the total width is \SI{20}{\micro\meter}. We pick periodic boundary conditions and drive the flow with a pressure gradient of \SI{-1e7}{\pascal\per\meter}. This simulation was done using the parameters obtained for the alginate based solution, as these represent the worst case scenario with respect to simulability among the fluids we work with. While our scheme is capable of simulating flows of to our knowledge arbitrary parameters, it increases runtime. For this parameter set the simulation time on our hardware becomes unreasonable. Therefore, for the simulations with the alginate based solutions we, contrary to all our other simulations, do not start from $0$ in all fields. We initialize the polymers in an extended state. This reduces the initial overshoot in the velocity by orders of magnitude allowing for larger time steps. Furthermore, the time to equilibrium of the polymer stretch is shortened considerably. This initialization leads to the simulation not accurately reproducing the time dependent behavior during startup, but leaves the steady state valid. The cross-section of the resulting flow field can be seen in figure \ref{fig:2DAlginateU}. Due to the long runtime, we stopped the simulation, early compared to the others. This means, that the simulation had not yet reached full equilibrium. Consequently, the error obtained here (see figure \ref{fig:2DAlginateErr}) is likely higher than what it would be after full relaxation. The average error as obtained by a typical L2 norm is \num{9e-4}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.eps}
		\caption{Plot of the cross-section of the velocity of an alginate channel flow. Likely not fully in equilibrium.}
		\label{fig:2DAlginateU}
	\end{figure}
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateErr.eps}
		\caption{Plot of error of the cross-section of the velocity of an alginate channel flow. Likely not fully in equilibrium.}
		\label{fig:2DAlginateErr}
	\end{figure}
	The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall. The same simulation is done for the 3D case in the following.

@@

	As can be seen in figure \ref{fig:malaspinas2010}, our results notably do not depend on the Weissenberg Number. While our error decreases with N with a constant power, Malaspinas et al. see diminishing returns. While these results are excellent, the error in our implementation is still dependent on the viscosity ratio. When going to significantly higher ratios, this needs to be checked to ensure it does not get too large.\todo{link to appendix where we validate this}
	With this, one can see, that we are not only capable to simulate parameters not yet accessible, but our implementation is also better for the parameters used in literature. Knowing our simulations to be good we in the following show results for interesting systems, that are not analytically solvable.
	\section{Results for non analytically solvable systems}\todo{do all of these with MC}

@@

	Conical nozzles are used for example in bio printing. For this use case in particular the stresses experienced within the nozzle are of importance. We simulate a cell with an inflow radius of \SI{1.8}{\milli\meter} and an outflow radius of \SI{200}{\micro\meter}. It has a length of \SI{18}{\milli\meter}. The boundary conditions for inflow and outflow are given as the result of a simulation of pipes of the respective radius with a fixed volume flow of \todo{find volume flow}. The resulting flow field can be seen in figure \ref{fig:Nozzle_v}. The von Mieses stress\todo{cite} can be seen in figure\ref{fig:Nozzle_sigma_vM}. It can be seen, that despite the high velocity in the center of the exit, the stress is actually smallest in the center. This means, that even for fast flows, given the bio ink and nozzle size is picked correctly, the cells flowing close to the center experience little stress.
	\todo{complete sim is wrong due to init issue}

@@

	A cell with a radius of \SI{6}{\micro\meter} and Youngs Modulus of \SI{100}{\pascal} is placed in a shear flow with a shear rate of \SI{400}{\per\second}. The fluid is described by the parameters of \SI{0.49}{\percent} methyl cellulose solution mentioned above. The lattice with a resolution of \num{100} lattice points is chosen significantly bigger than the cell to assure negligible impact of the walls. The lattice is periodic in $x$ and $z$ direction. At the edges of the simulation box in $y$ direction Dirichlet boundary conditions are imposed. The influence of the cell placed in the center of the simulation volume can barely seen in the velocity field (see figure \ref{fig:cellInShear_v}). In the von Mieses stress seen in figure \ref{fig:cellInShear_vM} it is the dominant feature. One can clearly see the shape of the cell by the depression of the stress it causes. Notable from this shape outwards radiate section of high and low stress. These are particularly pronounced along the longest axis of the deformed cell. From these one can see that the cell disturbs the fluid far from itself.
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/cellInShear_v.eps}
		\caption{Plot of the velocity field around a cell in shear flow.}
		\label{fig:cellInShear_v}

@@

	Real-time deformability cytometry (RT-DC)\todo{cite} is a novel approach for single cell characterization. The cells are forced through a constriction. The observed deformation can be used to infer the mechanical properties of the cell. For this simulation we use periodic boundary conditions and a pressure gradient\todo{specify}. The resulting flow can be seen in figure \ref{fig:RT-DC_v}. The component typically associated with viscosity of the stress tensor can be seen in figure \ref{fig:RT-DC_tau}. There are three qualitative differences compared to a purely viscous fluid. Directly after the inlet is a region of low stress. Due to the advection of the stress, the fluid here despite being inside the constriction has not reacted to it yet. This region therefore could produce unexpected cell behavior if one passes through it. Similarly, directly after the constriction, the stress has not yet lowered. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find an extremely low and an extremely high viscosity respectively. Depending on the exact parameters \glqq viscosities\grqq{ }can be reached that cannot be obtained through rheological measurements with the fluid. Meaning these viscosities are not valid. This is even clearer along the walls directly after the outlet. Here the viscosites obtained in this manner become negative. This shows the breakdown of the viscosity interpretation in viscoelastic cases and signifies the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by shear thinning models and should be considered despite the high computational cost.

>>>>

\title{Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids}
\author{Richard Kellnberger$^1$}
\date{January, 2024}
\begin{document}
	\maketitle

@@

	We need this solution to compare to literature. However, this solution does not include $\eta_\text{s}$, which is required in general. With the inclusion of the solvent viscosity an analytical solution can no longer be found and the use of a semi-analytic solution is required (for details see appendix \ref{app:PTT}).
	In the following we compare results for channel and pipe flow to these equations for realistic parameters. Afterward we compare the accuracy of our simulation algorithm against literature.
	%We use richard/vtk 4dbc4d379842c538f084e68a5dcfe2555a674495 (2024-01-09)
	\subsection{Channel flow vs theory}\label{sec:channelFlow}
	Channel flow (also known as 2D Poiseuille) is one of the simplest flows and a very typical example. We pick a channel with a radius of \SI{10}{\micro\meter}, meaning the total width is \SI{20}{\micro\meter}. We pick periodic boundary conditions and drive the flow with a pressure gradient of \SI{-1e7}{\pascal\per\meter}. This simulation was done using the parameters obtained for the alginate based solution, as these represent the worst case scenario with respect to simulability among the fluids we work with. While our scheme is capable of simulating flows of to our knowledge arbitrary parameters, it increases runtime. For this parameter set the simulation time on our hardware becomes unreasonable. Therefore, for the simulations with the alginate based solutions we, contrary to all our other simulations, do not start from $0$ in all fields. We initialize the polymers in an extended state. This reduces the initial overshoot in the velocity by orders of magnitude allowing for larger time steps. Furthermore, the time to equilibrium of the polymer stretch is shortened considerably. This initialization leads to the simulation not accurately reproducing the time dependent behavior during startup, but leaves the steady state valid. The cross-section of the resulting flow field can be seen in figure \ref{fig:2DAlginateU}. Due to the long runtime, we stopped the simulation, early compared to the others. This means, that the simulation had not yet reached full equilibrium. Consequently, the error obtained here (see figure \ref{fig:2DAlginateErr}) is likely higher than what it would be after full relaxation. The average error as obtained by a typical L2 norm is \num{9e-4}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.eps}
		\caption{Plot of the cross-section of the velocity of an alginate channel flow. Likely not fully in equilibrium.}
		\label{fig:2DAlginateU}
	\end{figure}
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateErr.eps}
		\caption{Plot of error of the cross-section of the velocity of an alginate channel flow. Likely not fully in equilibrium.}
		\label{fig:2DAlginateErr}
	\end{figure}
	The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall. The same simulation is done for the 3D case in the following.

@@

	As can be seen in figure \ref{fig:malaspinas2010}, our results notably do not depend on the Weissenberg Number. While our error decreases with N with a constant power, Malaspinas et al. see diminishing returns. While these results are excellent, the error in our implementation seemingly is still dependent on the viscosity ratio. As the error is still small using our alginate parameters (see section \ref{sec:channelFlow}), this effect is irrelevant. With this, one can see, that we are not only capable to simulate parameters not yet accessible, but our implementation is also better for the parameters used in literature. Knowing our simulations to be good we in the following show results for interesting systems, that are not analytically solvable.
	\section{Results for non analytically solvable systems}

@@

	Conical nozzles are used for example in bio printing. For this use case in particular the stresses experienced within the nozzle are of importance. We simulate a cell with an inflow radius of \SI{1.8}{\milli\meter} and an outflow radius of \SI{200}{\micro\meter}. It has a length of \SI{18}{\milli\meter}. The boundary conditions for inflow and outflow are given as the result of a simulation of pipes of the respective radius with a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$. The resulting flow field can be seen in figure \ref{fig:Nozzle_v}. The von Mieses stress\todo{cite} can be seen in figure\ref{fig:Nozzle_sigma_vM}. It can be seen, that despite the high velocity in the center of the exit, the stress is actually smallest in the center. This means, that even for fast flows, given the bio ink and nozzle size is picked correctly, the cells flowing close to the center experience little stress.

@@

	A cell with a radius of \SI{6}{\micro\meter} and Youngs Modulus of \SI{100}{\pascal} is placed in a shear flow with a shear rate of \SI{400}{\per\second}. The fluid is described by the parameters of \SI{0.49}{\percent} methyl cellulose solution mentioned above. The lattice with a resolution of \num{100} lattice points is chosen significantly bigger than the cell to assure negligible impact of the walls. The lattice is periodic in $x$ and $z$ direction. At the edges of the simulation box in $y$ direction Dirichlet boundary conditions are imposed. The influence of the cell placed in the center of the simulation volume can barely be seen in the velocity field. In the von Mieses stress seen in figure \ref{fig:cellInShear_vM} it is the dominant feature. One can clearly see the shape of the cell by the depression of the stress it causes. Notable from this shape outwards radiate section of high and low stress. These are particularly pronounced along the longest axis of the deformed cell. From these one can see that the cell disturbs the fluid far from itself.

@@

	Real-time deformability cytometry (RT-DC)\todo{cite} is a novel approach for single cell characterization. The cells are forced through a constriction. The observed deformation can be used to infer the mechanical properties of the cell. For this simulation we use periodic boundary conditions and a pressure gradient of \SI{1e7}{\pascal}. The resulting flow can be seen in figure \ref{fig:RT-DC_v}. The component typically associated with viscosity of the stress tensor can be seen in figure \ref{fig:RT-DC_tau}. There are three qualitative differences compared to a purely viscous fluid. Directly after the inlet is a region of low stress. Due to the advection of the stress, the fluid here despite being inside the constriction has not reacted to it yet. This region therefore could produce unexpected cell behavior if one passes through it. Similarly, directly after the constriction, the stress has not yet lowered. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find an extremely low and an extremely high viscosity respectively. Depending on the exact parameters \glqq viscosities\grqq{ }can be reached that cannot be obtained through rheological measurements with the fluid. Meaning these viscosities are not valid. This is even clearer along the walls directly after the outlet. Here the viscosites obtained in this manner become negative. This shows the breakdown of the viscosity interpretation in viscoelastic cases and signifies the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by shear thinning models and should be considered despite the high computational cost.
